\LARGE\singlespacing{\textbf{Supplementary Material} \\ \medskip Air cleaners and respiratory infections in schools: \\ A multi-measurement observational study}

### 2024-02-16 doc/comparison-mcid12/supplements.tex
Updates from Overleaf

\title{\LARGE\singlespacing{\textbf{Supplementary Material} \\ \medskip
Molecular detection of SARS-CoV-2 and other respiratory viruses in saliva, air, and filter samples from schools during and after the pandemic}}

@@

\begin{table}[!htpb]
    \caption{Differences in the study settings of 2022 and 2023.}
    \label{tab:comp_study}
    \centering
    \renewcommand{\arraystretch}{1.5}
    \begin{tabular}{lp{6cm}p{6cm}}
        \toprule
        & \textbf{2022 (during pandemic)} & \textbf{2023 (after pandemic)} \\
        \midrule
        Participants & 2~classes from School~1 sharing one classroom (half-class teaching) and 1~class from School~2 with a separate classroom & two classes from School~2 with separate classrooms \\
        Duration & 7 weeks per class (January 16 to March 11, 2023), excluding vacation & 7 weeks per class (January 24 to March 26, 2023), excluding vacation \\
        Saliva testing & weekly (Wednesdays) & bi-weekly (Tuesdays and Thursdays) \\
        Bioaerosol sampling & daily with BioSpot-VIVAS and Coriolis in both classrooms & daily with BioSpot-VIVAS and Coriolis in one classroom and only Coriolis in the other classroom \\
        HEPA-filters samples & collected once at the end of the study & collected twice after the phases where the air cleaner intervention was in place \\
        Interventions & compulsory face mask wearing as mandated by the public health authorities at that time from the start of the study for a total of 6 out of 14 study weeks; and installation of portable air cleaners towards the end of the study in the study classrooms for a total of 4 out of 14 study weeks & installation of portable air cleaners in the study classrooms using a cross-over design for a total of 7 out of 14 study weeks\\
        Ventilation conditions & all classrooms were not equipped with an active HVAC (heating, ventilation, air conditioning) system; passive window ventilation as per recommendations of the national public health authorities; in School~1 ventilation was additionally assisted by a CO$_2$ guided opener of a small window at the top  & all classrooms were not equipped with an active HVAC (heating, ventilation, air conditioning) system; passive window ventilation solely at the discretion of the teachers \\
        \bottomrule
    \end{tabular}
\end{table}

@@

\section{Model to estimate differences in airborne detection}\label{sec:model}

Our outcome is the binary variable $Y_{tcv}$ which is $1$ if any respiratory virus $v$ was detected in the air of classroom $c$ during study week $t$. We model this outcome with a Bayesian logistic regression model
\begin{align}
    \text{Bernoulli-Logit}(y|\mu) = \begin{cases}
        \text{logit}^{-1}(\mu) & \text{if }y=1\text{, and} \\
        1-\text{logit}^{-1}(\mu) & \text{if }y=0.
    \end{cases}
\end{align}
The parameter $\mu$ is related to the type of respiratory virus detected (SARS-CoV-2 vs non-SARS-CoV-2) and takes into account differences in the study setting (see \Cref{tab:comp_study}) as follows
\begin{align}
    \mu_{tcv} = \log(\text{offset}_c) + \alpha + \beta_1\,\cdot\text{saliva}_{tcv} + \beta_2\,\cdot\text{CoV}_{tcv} + \beta_3\,\cdot\text{mask}_{tcv} + \beta_4\,\cdot\text{aircl}_{tcv} + \beta_5\,\cdot\text{CO}_{2,tcv}.
\end{align}
Since one classroom in 2022 was shared by two classes, we set the \textbf{offset} of this classroom to 2 (double odds for airborne detection in this classroom). Conversely, since one classroom in 2023 only had one instead of two sampling devices in place, we set the \textbf{offset} of this classroom to 0.5 (half odds for airborne detection in this classroom). Other classrooms have an offset of 1. The binary variable \textbf{saliva} indicates if there was a positive saliva sample in the same week, considering with $\beta_1$ that airborne detection was more frequent if the same virus was already detected in saliva. The binary variable \textbf{CoV} indicates if the detected virus was SARS-CoV-2, thus our parameter of interest $\beta_2$ is the estimated log odds for detecting any SARS-CoV-2 versus non-SARS-CoV-2 in bioaerosols. The binary variables \textbf{masks} and \textbf{aircl} indicate if compulsory face mask wearing was mandated or the portable air cleaners were installed in the classroom, respectively, considering with $\beta_3$ and $\beta_4$ that these interventions possibly reduce the probability of airborne detection. The continuous variable \textbf{CO}$_2$ is the weekly average of the maximum daily CO$_2$ levels per classroom, considering with $\beta_5$ that higher CO$_2$ levels (a proxy for ventilation through greater outdoor air exchange) possibly increase the probability of airborne detection.

We model the effects of the aforementioned variables with a mix of  non-informative and informative priors. The intercept is modeled with a non-informative prior
\begin{align}
    \alpha \sim \text{Student}(\nu=5, \mu = 0, \sigma = 2.5) ~.
\end{align}
Since bioaerosols were more frequently detected in weeks with a positive saliva sample, we model $\beta_1$ with an informative prior
\begin{align}
    \beta_1 \sim \text{Normal}(\mu = 5/3, \sigma = s_y / s_{\text{saliva}}),
\end{align}
where 5 is the number of weeks with paired samples and 3 is the number of weeks with a positive bioaerosol but without a positive saliva sample. The prior's hyperparameter for the standard deviation is computed as the ratio of the standard deviation of the outcome $s_y$ and the variable $s_{\text{saliva}}$. \\
Our parameter of interest $\beta_2$ is modeled with a non-informative prior
\begin{align}
    \beta_2 \sim \text{Normal}(\mu = 0, \sigma = s_y / s_\text{CoV})~.
\end{align}
Since masks and air cleaners possibly reduce the odds of molecular airborne detection, we used an informative asymmetric Laplace prior for $\beta_3$ and $\beta_4$ as follows
\begin{align}
    \beta_3 &\sim \text{Asymmetric-Laplace}(\mu = -0.32, \sigma = 0.5\cdot s_y / s_{\text{masks}}, \tau = 0.75) \\
    \beta_4 &\sim \text{Asymmetric-Laplace}(\mu = -0.27, \sigma = 0.5\cdot s_y / s_{\text{aircl}}, \tau = 0.75),
\end{align}
which both place 90\% prior probability on negative intervention effects and are skewed towards smaller effect sizes. \\
Similarly, we incorporate our prior belief that higher CO$_2$ levels possibly increase the odds of airborne detection with the converse prior for $\beta_5$ as
\begin{align}
    \beta_3 &\sim \text{Asymmetric-Laplace}(\mu = 0.21, \sigma = 0.5\cdot s_y / s_{\text{masks}}, \tau = 0.25)~.
\end{align}

To summarize, we estimate the odd ratio for airborne detection of SARS-CoV-2, adjusting for differences in the setup of the classrooms (offset), positive saliva sample in the same week (saliva), the effects of interventions (masks and aircl), and differences in ventilation conditions (CO$_2$). We use informative priors for the adjustments because the sample size and between-years comparison otherwise make it difficult to identify their effects.

>>>>

\title{\LARGE\singlespacing{\textbf{Supplementary Material} \\ \medskip
Molecular detection of SARS-CoV-2 and other respiratory viruses in saliva and classroom air: a two winters tale}}

@@

\begin{table}[!htpb]
    \caption{Differences in the study settings of 2022 and 2023.}
    \label{tab:comp_study}
    \centering
    \renewcommand{\arraystretch}{1.5}
    \begin{tabular}{lp{6cm}p{6cm}}
        \toprule
        & \textbf{Winter 2021/22} & \textbf{Winter 2022/23} \\
        \midrule
        Participants & 2~classes from School~1 sharing one classroom (half-class teaching) and 1~class from School~2 with a separate classroom; School~1 and~2 are in different cities, which have grown together to form a community (\ie the schools are located nearby) & two classes from School~2 with separate (two) classrooms; none of the two classes already participated in winter 2021/22 \\
        Duration & 7 weeks per class (January 16 to March 11, 2023), excluding vacation & 7 weeks per class (January 24 to March 26, 2023), excluding vacation \\
        Saliva testing & weekly (Wednesdays) & bi-weekly (Tuesdays and Thursdays) \\
        Bioaerosol sampling & daily with BioSpot-VIVAS and Coriolis in both classrooms & daily with BioSpot-VIVAS and Coriolis in one classroom and only Coriolis in the other classroom \\
        HEPA-filters samples & collected once at the end of the study & collected twice after the phases where the air cleaner intervention was in place \\
        Interventions & compulsory face mask wearing as mandated by the public health authorities at that time from the start of the study, for a total of 6 out of 14 study weeks; and installation of portable air cleaners towards the end of the study in the study classrooms for a total of 4 out of 14 study weeks; in total, 10 study weeks with interventions & installation of portable air cleaners in the study classrooms using a cross-over design, for a total of 7 out of 14 study weeks; in total 7 study weeks with interventions\\
        Ventilation conditions & all classrooms were not equipped with an active HVAC (heating, ventilation, air conditioning) system; passive window ventilation as per recommendations of the national public health authorities; in School~1 ventilation was additionally assisted by a CO$_2$ guided opener of a small window at the top  & all classrooms were not equipped with an active HVAC (heating, ventilation, air conditioning) system; passive window ventilation solely at the discretion of the teachers \\
        \bottomrule
    \end{tabular}
\end{table}

@@

\begin{figure}
    \centering
    \includegraphics{}
    \caption{Caption}
    \label{fig:enter-label}
\end{figure}

\section*{Model to estimate differences in airborne detection}\label{sec:model}

Our outcome is the binary variable $Y_{tcv}$ which is $1$ if any respiratory virus $v$ was detected in the air of classroom $c$ during study week $t$. We model this outcome with a Bayesian logistic regression model
\begin{align}
    \text{Bernoulli-Logit}(y|\mu) = \begin{cases}
        \text{logit}^{-1}(\mu) & \text{if }y=1\text{, and} \\
        1-\text{logit}^{-1}(\mu) & \text{if }y=0.
    \end{cases}
\end{align}
The parameter $\mu$ is related to the type of respiratory virus detected (SARS-CoV-2 vs non-SARS-CoV-2) and takes into account differences in the study setting (see \Cref{tab:comp_study}) as follows
\begin{align}
    \mu_{tcv} = \log(\text{offset}_c) + \alpha + \beta_1\,\cdot\text{saliva}_{tcv} + \beta_2\,\cdot\text{CoV}_{tcv} + \beta_3\,\cdot\text{mask}_{tc} + \beta_4\,\cdot\text{aircl}_{tc} + \beta_5\,\cdot\text{CO}_{2,tc}.
\end{align}
Since one classroom in 2022 was shared by two classes, we set the \textbf{offset} of this classroom to 2 (double odds for airborne detection in this classroom). Conversely, since one classroom in 2023 only had one instead of two sampling devices in place, we set the \textbf{offset} of this classroom to 0.5 (half odds for airborne detection in this classroom). Other classrooms have an offset of 1. The binary variable \textbf{saliva} indicates if there was a positive saliva sample in the same week, with $\beta_1$ estimating whether airborne detection was more frequent if the same virus was already detected in saliva. This variable implicitly adjusts for community transmission, which would impact molecular detection both in saliva and airborne samples. The binary variable \textbf{CoV} indicates if the detected virus was SARS-CoV-2, thus our parameter of interest $\beta_2$ is estimating whether airborne detection of any SARS-CoV-2 is more likely than non-SARS-CoV-2 viruses. The binary variables \textbf{masks} and \textbf{aircl} indicate if compulsory face mask wearing was mandated or portable air cleaners were installed in the classroom, respectively, with $\beta_3$ and $\beta_4$ estimating the change in the probability of airborne detection due to the intervention. The continuous variable \textbf{CO}$_2$ is the weekly average of the maximum daily CO$_2$ levels per classroom, with $\beta_5$ estimating the impact of CO$_2$ levels (a proxy for classroom ventilation) on the probability of airborne detection.

Considering the relatively small sample size and prior knowledge, we model the effects of the aforementioned variables with a mix of  non-informative and informative priors. The intercept is modeled with a non-informative prior
\begin{align}
    \alpha \sim \text{Student}(\nu=5, \mu = 0, \sigma = 2.5) ~.
\end{align}
Since bioaerosols were more frequently detected in weeks with a positive saliva sample, we model $\beta_1$ with an informative prior
\begin{align}
    \beta_1 \sim \text{Normal}(\mu = 5/3, \sigma = s_y / s_{\text{saliva}}),
\end{align}
where 5 is the number of weeks with paired samples and 3 is the number of weeks with a positive bioaerosol but without a positive saliva sample. The prior's hyperparameter for the standard deviation is computed as the ratio of the standard deviation of the outcome $s_y$ and predictor $s_{\text{saliva}}$. \\
Our parameter of interest $\beta_2$ is modeled with a non-informative prior
\begin{align}
    \beta_2 \sim \text{Normal}(\mu = 0, \sigma = s_y / s_\text{CoV})~.
\end{align}
Since masks and air cleaners possibly reduce the odds of airborne detection, we used informative asymmetric Laplace priors for $\beta_3$ and $\beta_4$ as follows
\begin{align}
    \beta_3 &\sim \text{Asymmetric-Laplace}(\mu = -0.32, \sigma = 0.5\cdot s_y / s_{\text{masks}}, \tau = 0.75) \\
    \beta_4 &\sim \text{Asymmetric-Laplace}(\mu = -0.27, \sigma = 0.5\cdot s_y / s_{\text{aircl}}, \tau = 0.75),
\end{align}
which both place 90\% prior probability on negative intervention effects (decreasing the odds of airborne detection) and are skewed towards smaller effect sizes. \\
Similarly, we incorporate our prior belief that higher CO$_2$ levels possibly increase the odds of airborne detection with an informative asymmetric Laplace prior for $\beta_5$ as
\begin{align}
    \beta_3 &\sim \text{Asymmetric-Laplace}(\mu = 0.21, \sigma = 0.5\cdot s_y / s_{\text{masks}}, \tau = 0.25)~,
\end{align}
which places 90\% prior probability on positive effects (increasing the odds of airborne detection).

To summarize, we estimate the probability of airborne detection of SARS-CoV-2, adjusting for differences in the setup of the classrooms (offset), positive saliva sample in the same week (saliva), the effects of interventions (masks and aircl), and differences in ventilation conditions (CO$_2$). We use informative priors for the parameters of the adjustments to incorporate prior knowledge and because the sample size and between-years comparison would otherwise make it difficult to identify these parameters.

### 2024-02-16 doc/paper/supplements.tex
Updates from Overleaf

\section{Modeling positivity rate of human saliva samples}\label{sec:multinomial-model}

Molecular analysis determined which saliva samples were positive for a respiratory virus. Let $p = 0, 1, \dots, P$ denote the virus, where $0$ refers to negative samples and $P$ is the number of different viruses detected over the study. The number of positive bioaerosol and saliva samples $y_p$ in samples of sizes $n$ is analyzed with a Multinomial logistic regression model
\begin{align*}
    y_0, y_1, \cdots, y_P | n, \beta_0, \beta_1, \beta_2, \beta_3 &\sim \text{Multinomial-Logit}(\theta_0, \theta_1, \dots \theta_P) \\
    \theta &= \text{softmax}(\mu) \\
    \mu &= \begin{cases}
                0, & \text{if } p=0, \\
                \beta_{0p} + \beta_{1} \cdot \text{AirCleaner} + \beta_2 \cdot \text{Class} + \beta_3 \cdot \text{Susceptibles}_p, & \text{if } p>0.
            \end{cases} \\
    \beta_{0p} &\sim \text{Normal}(0, 2.5) \quad \forall p \\
    \beta_1, \beta_2, \beta_3 &\sim \text{Student-t}_5\left(0, \frac{2.5}{s_{x}}\right) 
\end{align*}
where $\text{softmax}(\mu) = \exp(\mu)/\sum\exp(\mu)$ and $s_{x}$ is the empirical standard deviation of each input variable.  The negative test is set as the reference category. Overall variation in positive samples by virus will be modeled with $\beta_{0p}$ and the effect of air cleaners will be modeled with $\beta_1$. The effect of air cleaners is adjusted for class-specific effects and the decreasing number of susceptibles over time (for each virus computed as the number of students minus the total number of positive samples).

Note that we deviated from the statistical analysis plan by estimating $\beta_{0p}$ and $\beta_1$ without a hierarchical prior. First, there was considerable variation in the overall positivity rate of each virus, so that we decided to use unpooled intercepts $\beta_{0p}$ instead of partially pooled intercepts $\beta_{0[p]}$ for each virus. Second, there was insufficient variation in the data to inform partially pooled estimates $\beta_{1[p]}$ for the effects of air cleaners. Therefore, we decided to only estimate the average effect across viruses, \ie the completely pooled estimate $\beta_1$.

@@

\section{Detailed results for positivity rate in saliva samples}
\label{sec:detailed-molecular}

There can be unknown delay between when students were infected and when they tested positive. Therefore, we performed a sensitivity analysis where we lagged the molecular test results (number of positive saliva samples) by one test date (\ie from Tuesday to last Thursday and from Thursday to Tuesday) and one test week (\ie from Tuesday to last Tuesday and Thursday to last Thursday). The results are shown in \Cref{fig:mol-estimation-results-sensitivity}. The estimated effect of air cleaners remains indistinguishable from zero regardless of the considered lag.  
\begin{figure}[!htb]
\centering
    \includegraphics[width=10cm]{../../results/mol-data/model-results.pdf}
    \caption[Sensitivity analysis for the estimated effect of air cleaners on the positivity rate of saliva samples]{Sensitivity analysis for the estimated effect of air cleaners on the positivity rate of saliva samples. Shown is the adjusted relative risk (posterior mean as dot and 50\%-, 80\%, and 95\%-CrIs as lines) for air cleaners by lag of the molecular test results.}
    \label{fig:mol-estimation-results-sensitivity}
\end{figure}

>>>>

% \section{Modeling positivity rate of human saliva samples}\label{sec:multinomial-model}

% Molecular analysis determined which saliva samples were positive for a respiratory virus. Let $p = 0, 1, \dots, P$ denote the virus, where $0$ refers to negative samples and $P$ is the number of different viruses detected over the study. The number of positive bioaerosol and saliva samples $y_p$ in samples of sizes $n$ is analyzed with a Multinomial logistic regression model
% \begin{align*}
%     y_0, y_1, \cdots, y_P | n, \beta_0, \beta_1, \beta_2, \beta_3 &\sim \text{Multinomial-Logit}(\theta_0, \theta_1, \dots \theta_P) \\
%     \theta &= \text{softmax}(\mu) \\
%     \mu &= \begin{cases}
%                 0, & \text{if } p=0, \\
%                 \beta_{0p} + \beta_{1} \cdot \text{AirCleaner} + \beta_2 \cdot \text{Class} + \beta_3 \cdot \text{Susceptibles}_p, & \text{if } p>0.
%             \end{cases} \\
%     \beta_{0p} &\sim \text{Normal}(0, 2.5) \quad \forall p \\
%     \beta_1, \beta_2, \beta_3 &\sim \text{Student-t}_5\left(0, \frac{2.5}{s_{x}}\right) 
% \end{align*}
% where $\text{softmax}(\mu) = \exp(\mu)/\sum\exp(\mu)$ and $s_{x}$ is the empirical standard deviation of each input variable.  The negative test is set as the reference category. Overall variation in positive samples by virus will be modeled with $\beta_{0p}$ and the effect of air cleaners will be modeled with $\beta_1$. The effect of air cleaners is adjusted for class-specific effects and the decreasing number of susceptibles over time (for each virus computed as the number of students minus the total number of positive samples).

% Note that we deviated from the statistical analysis plan by estimating $\beta_{0p}$ and $\beta_1$ without a hierarchical prior. First, there was considerable variation in the overall positivity rate of each virus, so that we decided to use unpooled intercepts $\beta_{0p}$ instead of partially pooled intercepts $\beta_{0[p]}$ for each virus. Second, there was insufficient variation in the data to inform partially pooled estimates $\beta_{1[p]}$ for the effects of air cleaners. Therefore, we decided to only estimate the average effect across viruses, \ie the completely pooled estimate $\beta_1$.

@@

% \section{Detailed results for positivity rate in saliva samples}
% \label{sec:detailed-molecular}

% There can be unknown delay between when students were infected and when they tested positive. Therefore, we performed a sensitivity analysis where we lagged the molecular test results (number of positive saliva samples) by one test date (\ie from Tuesday to last Thursday and from Thursday to Tuesday) and one test week (\ie from Tuesday to last Tuesday and Thursday to last Thursday). The results are shown in \Cref{fig:mol-estimation-results-sensitivity}. The estimated effect of air cleaners remains indistinguishable from zero regardless of the considered lag.  
% \begin{figure}[!htb]
% \centering
%     \includegraphics[width=10cm]{../../results/mol-data/model-results.pdf}
%     \caption[Sensitivity analysis for the estimated effect of air cleaners on the positivity rate of saliva samples]{Sensitivity analysis for the estimated effect of air cleaners on the positivity rate of saliva samples. Shown is the adjusted relative risk (posterior mean as dot and 50\%-, 80\%, and 95\%-CrIs as lines) for air cleaners by lag of the molecular test results.}
%     \label{fig:mol-estimation-results-sensitivity}
% \end{figure}

\section{Possible transmission networks}\label{sec:transmission-networks}

The distribution of positive saliva samples varied between classes. For example, all but one sample was positive for adenvovirus in class~A during the first three study weeks, while the vast majority of positive influenza~B samples were found in class~B. To illustrate possible transmission chains within classes, we linked positive saliva samples of the same virus that were less than one week apart. Based on this spatiotemporal analysis, we identified 10~possible transmission chains (\Cref{fig:molecular-descriptives}). The longest potential transmission chains occurred in January, referring to a cluster of adenovirus infections in class~A and influenza~B infections in class~B. Molecular genotyping to verify the proposed transmission network was unsuccessful because we could not amplify and sequence any of the gene targets. We also analyzed 105~bioaerosol samples and detected in two of them viral RNA (1~rhinovirus in class~A and 1~adenovirus in class~B). Similarly, we detected 1~influenza~B, 1~rhinovirus, 1~adenovirus, and 1~SARS-CoV-2 in the 20~swabs taken from each filter of an air cleaner after each intervention phase (160~swabs in total).

\begin{figure}[!htpb]
    \centering
    \includegraphics{../../results/mol-data/saliva-distribution.pdf}\hspace{.5cm}
    \includegraphics{../../results/mol-data/network-plot.pdf}
    \caption{\textbf{Molecular detection of respiratory viruses and transmission network based on spatiotemporal analysis of students' saliva samples}. \textbf{(a)}~Number of positive saliva samples by virus. \textbf{(b)}~Daily number of positive saliva samples (colored circles) and possible transmission chains within classes (directed arrows). Positive samples are linked if they belong to the same virus and are less than 1~week apart. Positive samples from the air and filters as blank squares aligned. IFB: influenza~B, HRV: human rhinovirus, AdV: adenovirus, CoV: SARS-CoV-2, MPV: human metapneumovirus, PIV: parainfluenza virus.}
    \label{fig:molecular-descriptives}
\end{figure}
